Revisiting the hydrodynamic modulation of short surface waves by longer waves

### 2024-10-12 paper.tex
Fix caption
Expand on limitations

\begin{equation}
\label{eq:lhs1960wavenumber}
\widetilde{k} = k (1 + \varepsilon_L)
\end{equation}
where $\widetilde{k}$ and $k$ are the modulated and unmodulated short-wave wavenumbers,
respectively, and $\varepsilon_L = a_L k_L$ is the steepness of the long wave with
wavenumber $k_L$ and amplitude $a_L$.
This process is illustrated in Fig. \ref{fig:hydrodynamic_modulation_diagram}.
Hydrodynamic modulation is a key process for the development and
interpretation of remote sensing of ocean surface waves and currents
\citep{keller1975microwave,hara1994hydrodynamic}, and is distinguished
from aerodynamic modulation \citep{donelan1987effect,belcher1999wave,chen2000effects}.
\citet{phillips1981dispersion} and \citet{longuet1987propagation} revisited the
hydrodynamic modulation problem by considering nonlinear long waves and
variations in the effective gravity of short waves.
Their results yielded significantly stronger modulation than previously predicted.
\citet{henyey1988energy} derived an analytical solution for the modulation of
short waves using Hamiltonian mechanics and reported similar modulation magnitudes
to that of \citet{longuet1987propagation}.
\citet{zhang1990evolution} considered weakly nonlinear short waves
using a non-linear Schr\" odinger equation and found stronger wavenumber
modulation but weaker amplitude modulation than those predicted by
\citet{longuet1987propagation}.
The variety of analytical and numerical frameworks (Eulerian crest and action
conservation, Hamiltonian, Schr\" odinger) that relatively closely reproduce
steady solutions for short-wave modulation suggests that the solutions are robust.
Nevertheless, alternative analytical and numerical approaches to the modulation
problem remain of interest, especially if higher degrees of nonlinearity can be
considered using simpler approaches.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/fig_hydrodynamic_modulation_diagram.pdf}
\caption{
  A diagram of short waves riding on longer waves that propagate with their
  phase-speed $C_{pL}$ from left to right.
  The short waves propagate on the long-wave surface with their group velocity
  $C_g$.
  Long waves induce orbital velocities $U$ that cause near-surface convergence
  and divergence on their front and rear faces, respectively.
  They also induce downward and upward centripetal accelerations
  ($\partial W/\partial t$) in the crests and troughs, respectively, that
  modulate the effective gravity at the surface.
  As a consequence of surface convergence and divergence and the effective-gravity
  modulation, the short waves become shorter and higher preferentially
  on the long-wave crests.
  The diagram is not to scale.
}
\label{fig:hydrodynamic_modulation_diagram}
\end{figure}

@@

In this paper, we revisit this problem from linear wave theory but consider
the long-wave slope to be significant enough to require evaluating the
short-waves at the long-wave surface rather than at the mean water level.
This allows us to derive alternative, steady, non-linear solutions for the
modulation of short waves.
The analytical solutions yield modulations that are stronger than the original
solution by \citet{longuet1960changes}, but similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution} in
long-wave steepness up to $\approx 0.2$.
To validate the steadiness of the analytical solutions, we perform numerical
simulations of the full wave crest and action conservation laws, and compare
the results to the analytical solutions presented here, as well as the prior
numerical results.
The numerical simulations are used to determine the validity of the approximate
steady solution and to investigate the unsteadiness of the short-wave modulation
found by \citet{peureux2021unsteady}.
The main prior studies on hydrodynamic modulation, and their assumptions and
approaches to the solution, are summarised in Table
\ref{table:modulation_literature_summary}.
We first present the governing equations in \S\ref{section:governing_equations},
and then the analytical solutions for the modulation of short waves by long waves
in \S\ref{section:analytical_solutions}.
Then, we describe the numerical simulations and their results in
\S\ref{section:numerical_solutions}.
Finally, we discuss the results and their implications in
\S\ref{section:discussion} and conclude the paper in \S\ref{section:conclusions}.

\begin{table}
\begin{center}
\def~{\hphantom{0}}
\begin{tabular}{cccccc}
Reference   & Governing & Gravity & Nonlinear  & Nonlinear   & Solution \\
            & equations & varies  & long waves & short waves &          \\
\hline
\citet{longuet1960changes} & Velocity potential  & No  & No  & No  & Analytical \\
\hline
\citet{phillips1981dispersion} & Crest and action & Yes & Yes & No  & Numerical \\
                               & conservation     &     &     &     &           \\
\hline
\citet{longuet1987propagation} & Velocity potential & Yes & Yes & No  & Numerical \\
\hline
\citet{henyey1988energy} & Hamiltonian & Yes & Yes & No  & Analytical \\
\hline
\citet{zhang1990evolution} & Schr\" odinger & Yes & Yes & Yes & Numerical \\
\hline
\citet{peureux2021unsteady} & Crest and action & Yes & No  & No  & Numerical \\
                            & conservation     &     &     &     &           \\
\hline
This paper  & Crest and action & Yes & Yes & No  & Analytical \\
            & conservation     &     &     &     & and numerical \\
\hline
\end{tabular}
\caption{
  Summary of prior and present (this paper) approaches to calculating the
  modulation of short waves by long waves.
}
\label{table:modulation_literature_summary}
\end{center}
\end{table}

@@

(\ref{eq:dispersion}) and (\ref{eq:wave_crests}) only describe the evolution of
the short-wave wavenumber.
To describe the evolution of the wave amplitude, use the wave action balance
\citep{bretherton1968wavetrains} in absence of sources (e.g. due to wind) and
sinks (e.g. due to whitecapping):

\begin{equation}
\label{eq:wave_action}
\dfrac{\partial N}{\partial t}
+ \dfrac{\partial}{\partial x} \left[\left(C_g + U\right)N\right]
= 0
\end{equation}
where $N$ is the action density, and $C_g = 1/2\sqrt{g/k}$ is the group speed
in deep water.
Although the absence of wave growth and dissipation is not generally realistic
in the open ocean, it is a useful approximation here to isolate the effects
of hydrodynamic modulation solely due to the motion of the long waves.

@@

For the ambient forcing, we consider a train of monochromatic long wave
defined, to the first order in steepness, by the elevation $\eta_L = a_L \cos{\psi}$,
where $\psi = k_L x - \sigma_L t$ is the long-wave phase and $\sigma_L$ its
angular frequency.
The long wave induces horizontal and vertical orbital velocities whose values
at the surface $\eta_L$ are:

@@

\begin{equation}
\label{eq:w_L}
W = a_L \omega_L e^{\varepsilon_L \cos{\psi}} \sin{\psi}
\end{equation}

(\ref{eq:dispersion})-(\ref{eq:w_L}) are the governing equations used in this paper.
Note that rather than deriving the modulation solution from the velocity
potential of two waves, as \citet{longuet1960changes} did, starting from the
crest-action conservation balances allows for a simpler derivation.

@@

Although in this paper we only explored the solutions for $k_L = 1$ and $k=10$
the results are not sensitive to different values of $k_L$, provided that the
amplitude $a_L$ is set such that the long-wave steepness $\varepsilon_L = a_L / k_L$
is the same.
Numerical simulations with $k/k_L = 10$ and $k/k_L = 100$ show little
sensitivity to the ratio $k/k_L$ (not shown), consistent with the results of
\citet{longuet1987propagation}.
Specifically, the short-wave steepness modulation is 1\% larger at
$\varepsilon_L = 0.27$ for $k/k_L = 10$ compared to $k/k_L = 100$, and 4\%
larger at $\varepsilon_L = 0.4$.
This suggests that as long as $k/k_L \gg 1$, the modulation amplitude is
largely independent of the ratio $k/k_L$.
These results can be reproduced with the companion numerical model.

>>>>

\begin{equation}
\label{eq:lhs1960wavenumber}
\widetilde{k} = k (1 + \varepsilon_L \cos{\psi})
\end{equation}
where $\widetilde{k}$ and $k$ are the modulated and unmodulated short-wave wavenumbers,
respectively, and $\varepsilon_L = a_L k_L$ is the steepness of the long wave with
wavenumber $k_L$ and amplitude $a_L$, and $\psi$ is the long-wave phase.
This process is illustrated in Fig. \ref{fig:hydrodynamic_modulation_diagram}.
Hydrodynamic modulation is a key process for the development and
interpretation of remote sensing of ocean surface waves and currents
\citep{keller1975microwave,hara1994hydrodynamic}, and is distinguished
from aerodynamic modulation \citep{donelan1987effect,belcher1999wave,chen2000effects}.
\citet{phillips1981dispersion} and \citet{longuet1987propagation} revisited the
hydrodynamic modulation problem by considering nonlinear long waves and
variations in the effective gravity of short waves.
Their results yielded significantly stronger modulation than previously predicted.
\citet{henyey1988energy} derived an analytical solution for the modulation of
short waves using Hamiltonian mechanics and reported similar modulation magnitudes
to that of \citet{longuet1987propagation}.
\citet{zhang1990evolution} considered weakly nonlinear short waves
using a nonlinear Schr\" odinger equation and found stronger wavenumber
modulation but weaker amplitude modulation than those predicted by
\citet{longuet1987propagation}.
The variety of analytical and numerical frameworks (Eulerian crest and action
conservation, Hamiltonian, Schr\" odinger) that relatively closely reproduce
steady solutions for short-wave modulation suggests that the solutions are robust.
Nevertheless, alternative analytical and numerical approaches to the modulation
problem remain of interest, especially if higher degrees of nonlinearity can be
considered using simpler approaches.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/fig_hydrodynamic_modulation_diagram.pdf}
\caption{
  A diagram of short waves riding on longer waves that propagate with their
  phase-speed $C_{pL}$ from left to right.
  The short waves propagate on the long-wave surface, their action density
  moving with the group velocity $C_g$.
  Long waves induce orbital velocities $U$ that cause near-surface convergence
  and divergence on their front and rear faces, respectively.
  They also induce downward and upward centripetal accelerations
  ($\partial W/\partial t$) in the crests and troughs, respectively, that
  modulate the effective gravity at the surface.
  As a consequence of surface convergence and divergence and the effective-gravity
  modulation, the short waves become shorter and higher preferentially
  on the long-wave crests.
  The diagram is not to scale.
}
\label{fig:hydrodynamic_modulation_diagram}
\end{figure}

@@

In this paper, we revisit this problem from linear wave theory but consider
the long-wave slope to be significant enough to require evaluating the
short-waves at the long-wave surface rather than at the mean water level.
This allows us to derive alternative, steady, nonlinear solutions for the
modulation of short waves.
The analytical solutions yield modulations that are stronger than the original
solution by \citet{longuet1960changes}, but similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution} in
long-wave steepness up to $\approx 0.2$.
To validate the steadiness of the analytical solutions, we perform numerical
simulations of the full wave crest and action conservation laws, and compare
the results to the analytical solutions presented here, as well as the prior
numerical results.
The numerical simulations are used to determine the validity of the approximate
steady solution and to investigate the unsteadiness of the short-wave modulation
found by \citet{peureux2021unsteady}.
The main prior studies on hydrodynamic modulation, and their assumptions and
approaches to the solution, are summarised in Table
\ref{table:modulation_literature_summary}.
We first present the governing equations in \S\ref{section:governing_equations},
and then the analytical solutions for the modulation of short waves by long waves
in \S\ref{section:analytical_solutions}.
Then, we describe the numerical simulations and their results in
\S\ref{section:numerical_solutions}.
Finally, we discuss the results and their implications in
\S\ref{section:discussion} and conclude the paper in \S\ref{section:conclusions}.

\begin{table}
\begin{center}
\def~{\hphantom{0}}
\begin{tabular}{cccccc}
Reference   & Analysis   & Gravity & Nonlinear  & Nonlinear   & Solution \\
            & frameworks & varies  & long waves & short waves &          \\
\hline
\citet{longuet1960changes} & Velocity potential  & No  & No  & No  & Analytical \\
\hline
\citet{phillips1981dispersion} & Crest and action & Yes & Yes & No  & Numerical \\
                               & conservation     &     &     &     &           \\
\hline
\citet{longuet1987propagation} & Velocity potential & Yes & Yes & No  & Numerical \\
\hline
\citet{henyey1988energy} & Hamiltonian & Yes & Yes & No  & Analytical \\
\hline
\citet{zhang1990evolution} & Schr\" odinger & Yes & Yes & Yes & Numerical \\
\hline
\citet{peureux2021unsteady} & Crest and action & Yes & No  & No  & Numerical \\
                            & conservation     &     &     &     &           \\
\hline
This paper  & Crest and action & Yes & Yes & No  & Analytical \\
            & conservation     &     &     &     & and numerical \\
\hline
\end{tabular}
\caption{
  Non-exhaustive summary of prior and present (this paper) approaches to
  calculating the modulation of short waves by long waves.
}
\label{table:modulation_literature_summary}
\end{center}
\end{table}

@@

(\ref{eq:dispersion}) and (\ref{eq:wave_crests}) describe the evolution of
the short-wave wavenumber.
To describe the evolution of the wave amplitude, use the wave action balance
\citep{bretherton1968wavetrains} in absence of sources (e.g. due to wind) and
sinks (e.g. due to whitecapping):

\begin{equation}
\label{eq:wave_action}
\dfrac{\partial N}{\partial t}
+ \dfrac{\partial}{\partial x} \left[\left(C_g + U\right)N\right]
= 0
\end{equation}
where $N$ is the action density of short waves defined as the ratio of their
energy density to their intrinsic frequency, $E/\sigma$,
and $C_g = 1/2\sqrt{g/k}$ is their group speed in deep water.
The wave action density expressed as $E/\sigma$ is commonly taken as
conservative in wave modulation theory
\citep{phillips1981dispersion,longuet1987propagation}, however as
\citet{dysthe1988orbiting} show, it is conservative only for gently sloped waves.
Although the absence of wave growth and dissipation is not generally realistic in
the open ocean, it is a useful approximation here to isolate the effects
of hydrodynamic modulation solely due to the motion of the long waves.

@@

For the ambient forcing, we consider a train of monochromatic long waves
defined, to the first order in $\varepsilon_L$, by the elevation
$\eta_L = a_L \cos{\psi}$, where $\psi = k_L x - \sigma_L t$ is the long-wave
phase and $\sigma_L$ its angular frequency.
The velocity potential of the long wave is:

\begin{equation}
\label{eq:phi_L}
\phi_L = \dfrac{a_L \omega_L}{k_L} e^{k_L z} \sin{\psi} + \mathcal{O}(\varepsilon_L^4)
\end{equation}
so, up to the third order in $\varepsilon_L$, the long-wave induced horizontal
and vertical orbital velocities, evaluated at the wave surface $z = \eta_L$, are:

@@

\begin{equation}
\label{eq:w_L}
W = a_L \omega_L e^{\varepsilon_L \cos{\psi}} \sin{\psi}
\end{equation}
Evaluating these velocities at the wave surface $z = \eta_L$ rather than at
the mean water level $z=0$ is a departure from the linear theory, making the
treatment of the long waves here as weakly nonlinear.

(\ref{eq:wavenumber})-(\ref{eq:wave_action2}) are the governing equations used
in this paper, with (\ref{eq:U_L})-(\ref{eq:w_L}) describing the long-wave
induced ambient velocity forcing for the short waves. 
Rather than deriving the modulation solution from the velocity
potential of two waves, as \citet{longuet1960changes} did, starting from the
crest-action conservation balances allows for a simpler derivation.

@@

Although in this paper we only explored the solutions for $k_L = 1$ and $k=10$
the results are not sensitive to different values of $k_L$, provided that the
amplitude $a_L$ is set such that the long-wave steepness $\varepsilon_L = a_L / k_L$
is the same.
Numerical simulations with $k/k_L = 10$ and $k/k_L = 100$ show little
sensitivity to the ratio $k/k_L$ (not shown), consistent with the results of
\citet{longuet1987propagation}.
Specifically, the short-wave steepness modulation is 1\% larger at
$\varepsilon_L = 0.27$ for $k/k_L = 10$ compared to $k/k_L = 100$, and 4\%
larger at $\varepsilon_L = 0.4$.
This suggests that as long as $k/k_L \gg 1$, the modulation amplitude is
largely independent of the ratio $k/k_L$.
These results can be reproduced with the companion numerical model.
Also not considered here are the short-wave trains that are oblique to the
longer waves, which is relatively straightforward to include
\citep{peureux2021unsteady}.
